Fast Sorting on Modern Computers

### 2005-07-26 docs/tech_report/ordernsquared.tex
renamed tech_report_docs to tech_report
minor changes to text started lists of results and contributions

\label{ordernsquared}
Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
manner in which they check every key in the array in order to sort a single key;
for every key to be sorted, every other key is checked.  Four of these sorts are
discussed here. They are insertion sort, selection sort, bubblesort and
shakersort.

@@

%TODO look at knuths number here
TODO Knuth has a formula which I will look up.
\section{Insertion sort}

@@

\code{Bubblesort2 code}{on2_bubblesort2.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the bubblesort2 should perform faster than the original in
all metrics. Instruction count, were there no conditions that allowed keys to be
skipped, would be the same, as no extra instructions are added. The flow of the
algorithm is the same, though there should be fewer iterations, especially at
the start, when iterations are longer, so cache accesses, and branch misses
should be reduced along with instruction count.

@@

Finally, the branch prediction results are expected to be the same as in bubble
sort, with one difference.  Keys which must travel the entire way across the
array are taken care of earlier in the sort. If the largest key is in the
leftmost position, it must be moved the entire way right, which causes misses
for every move in bubblesort, since each move is disjoint from the next. In
shakersort, however, each in the correct direction will follow a move in the
same direction, each of which should be predicted. This works for a lot of
keys, which are more likely to move in continuous movements than they are in
bubblesort, where only small keys on the right move continuously. The
performance should improve noticeably as a result.

@@

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less. The improvement to
bubblesort in bubblesort2 have increased the instruction count, contrary to
predictions, and similarly in the shakersort2.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse (see Section \ref{temporal locality}),
reducing its number of misses. Shakersort2 is placed roughly between insertion
sort and shakersort, showing that ending early is helpful in its case. This does
not appear to be the case with bubblesort2 though, which performs identically to
bubblesort.

>>>>

\label{ordernsquared}
Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
manner in which they check every key in the array in order to sort a single key;
for every key to be sorted, every other key is checked.  Four of these sorts are
discussed here: insertion sort, selection sort, bubblesort and shakersort,
along with improved versions of bubblesort and shakersort.

@@

%TODO look at knuths number here
TODO Knuth has a formula which I will look up.

\section{Insertion sort}

@@

\code{Bubblesort2 code}{on2_bubblesort2.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the bubblesort2 should perform faster than the original in
all metrics. Instruction count, were there no conditions that allowed keys to be
skipped, would be the similar, since few extra instructions are added. The flow
of the algorithm is the same, though there should be fewer iterations,
especially at the start, when iterations are longer, so cache accesses, and
branch misses should be reduced along with instruction count.

@@

Finally, the branch prediction results are expected to be the same as in bubble
sort, with one difference.  Keys which must travel the entire way across the
array are taken care of earlier in the sort. If the largest key is in the
leftmost position, it must be moved the entire way right, which causes misses
for every move in bubblesort, since each move is disjoint from the next. In
shakersort, however, each move in the correct direction will follow a move in the
same direction, each of which should be correctly predicted. This works for a
lot of keys, which are more likely to move in continuous movements than they are
in bubblesort, where only small keys on the right move continuously. The
performance should improve noticeably as a result.

@@

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less. The improvement to
bubblesort in bubblesort2 have increased the instruction count, contrary to
predictions, and similarly in the shakersort2. It appears the cost of
skipping over a segment of the array is not worth the additional book-keeping
cost of determining when this can be done.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse, reducing its number of misses. Shakersort2
is placed roughly between insertion sort and shakersort, showing that ending
early is helpful in its case. This does not appear to be the case with
bubblesort2 though, which performs identically to bubblesort.
